Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest

### 2022-07-11 article/abs.Rmd
first poster draft

About half of all forests are tropical and secondary, making tropical forest regeneration integral to future forests. 
Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely. 
Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration, mediated by shade tolerance and dispersal traits, along with management history.
This study addresses how wet tropical secondary forest regeneration is affected by distance to habitat edge, and hypothesizes that old timber plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and adjacent forest edges mediate community composition.
An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to traits using relevant literature.

<!-- # Results/Conclusions -->

As distance from primary forest edge increased, stem and wood density tended to increase significantly, with ca 10% variation explained, while biomass and canopy light tended to stay the same.
Stand tree richness also tended to increase significantly, but diversity decreased steeply and non-linearly, and taxonomic composition varied notably. 
Finally, tree taxa associated with both early and late successional stages decreased significantly, but biomass by dispersal mode did not tend to change.
Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.

>>>>

About half of all forests are tropical and secondary, making tropical forest regeneration integral to future forests. 
Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely. 
Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration, mediated by shade tolerance and dispersal traits, along with management history.
This study addressed how wet tropical secondary forest regeneration was affected by distance to habitat edge, and hypothesized that old timber plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and adjacent forest edges mediate community composition.
An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to traits using relevant literature.

<!-- # Results/Conclusions -->
As distance from primary forest edge increased, stem and wood density tended to increase significantly, with ca 10% variation explained, while biomass and canopy light tended to stay the same.
Stand tree richness also tended to increase significantly, but diversity decreased steeply and non-linearly, and taxonomic composition varied notably. 
Finally, tree taxa associated with both early and late successional stages decreased significantly, but biomass by dispersal mode did not tend to change.
Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.

\newpage

# Abstracto{-}

Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
Biomasa arboreal y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
Dado que la mayoría de bosques quedan entre un km de su orilla, efectos de borde probablemente afectan la regeneración de composición de comunidad, mediado por tolerancia a sombra y characterísticas de disperso, y también historia de manejo.
Este estudio investigó cómo la regeneración de bosque secundario tropical húmedo fue afectado por distancia a borde de hábitat, e hipotetizó que plantaciones de madera viejas facilitan regeneración porque favorecen a taxa tolerantes a sombra o de sucesión tarde, y que orillas de bosque adyacente median cambios a composición de comunidad.
Una plantación de madera neotropical adyacente a bosque primario fue censada por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combinó taxa con characterísticas ecológicas usando literatura relevante.

Mientras distancia de bosque primario aumentó, densidad de madera y tallos tendieron a aumentar significativamente, con 10% variación explicado, mientras tanto biomasa y luz de dócel tendió a quedarse igual.
Riquesa taxonómica de árboles también tendió a aumentar significativamente, pero diversidad bajó profundamente y no-linearmente, y composición taxonomica varió notablemente.
Finalmente, biomasa de taxa de árboles asociados con etapas de sucesión ambos temprano y tarde bajaron significativamente, pero biomasa por modo de disperso no tendió a cambiar.
Entre todo este estudio apoya que composición taxonómica es menos resiliente y más sujeto a efectos de borde negativos que biomasa y riqueza taxonómica, sugeriendo que bosques globales probablemente serán ensamblajes nuevamente distintos en el futuro, con compensaciones entre producción de madera y biodiversidad que ocurren basado en manejo local y regional.

### 2022-07-11 article/disc.Rmd
first poster draft

---
title: "Discussion"
output: html_notebook
bibliography: ../../library.bib
link-citations: TRUE
---

1.  assess hypotheses
2.  infer causes of patterns - plantation structure, old legacy trees, or species traits?
3.  compare inferences to other studies
4.  recommend future conservation approaches, ecological studies

Under more sustainable forestry that maintains multiple functions, forest stand regeneration is likely more predictable, generally mirroring (i.e. near 1:1 replacement) the targeted tree cutting that is done using information collected about the ecological profiles and ages (when possible) of trees (Duah-Gyamfi et al 2014, Edwards et al 2014).

In contrast, under intensive silviculture that prioritizes the growth of select tree crop species, forest stands can show high wood production (Ferez et al 2015) while also usually showing negative trade-offs with other services, like that with biodiversity given reliance on monoculture planting designs.

In some cases where management decisions are made primarily based on regional economic but not local ecological information, due to limited access to funded scientific reporting (Cook et al 2012), forest stands can perform very poorly all-around.

After this more extreme type of management, it remains uncertain how natural regeneration in secondary forests will proceed, how local ecosystem services will be restored, and how restoration could be managed locally (Laurance et al 2012), e.g. based on surrounding forest proximity.

The persistence of early-successional tree communities has been found to lower the standing biomass, likely due to lower wood density *[@césar21]*, and could therefore require a switch in focal timber species *[@karsten14]*.

>>>>

---
title: "Discussion"
output: 
  # html_notebook
  bookdown::html_document2:
    # number_sections: FALSE
bibliography: ../../library.bib
link-citations: TRUE
---

<!-- # outline{-} -->
<!-- 1.  assess hypotheses -->
<!-- 2.  infer causes of patterns - plantation structure, old legacy trees, or species traits? -->
<!-- 3.  compare inferences to other studies -->
<!-- 4.  recommend future conservation approaches, ecological studies -->

<!-- # draft{-} -->

This study found significant edge effects on secondary forest stand wood and stem density, tree diversity and community composition, and some functional traits, yet not on overall aboveground biomass.
<!-- These results were shown in Figs \@ref(fig:biomass), \@ref(fig:taxa), and \@ref(fig:traits)... -->
It was initially hypothesized that community composition would vary with distance to edge, favoring more shade-tolerant taxa, due to light availability and dispersal potential, and this evidence supported edge effects on community composition and biomass of taxa associated with both early- and late-successional stages, but without detectable roles for light availability or general dispersal mode.
Accordingly this suggests that edge effects do significantly change humid tropical forest community composition, but primarily due to factors other than light competition or dispersal limitation, as well as that aboveground biomass and even low taxonomic richness is resilient after a decade of regeneration.

Aboveground biomass storage is a key global ecosystem function and service, but edge effects tend to release about one-third of carbon lost from tropical deforestation *[@brinck17]*, with more expected in the future *[@mitchard18]*.
<!-- responds variably to secondary succession and edge effects ...   -->
Management has been found to explain most of biomass variation among tropical forest fragments, and wood density ~10% biomass variation *[@pyles22]*.
This study adds that management via edge effects can also mediate ~10% of wood density variation without affecting local biomass, although median stem wood density can describe regional spatial biomass patterns *[@baker04]*.
Overall, wood density and richness values may have been pulled down by early-successional clonal trees closest to the primary forest, which tend to have lower wood densities, lowering carbon *[@gonzalezj98; @pyles22]*, and increasing nearby stem density.
Lower wood density can also reflect more dynamic forest stands *[@malhi06]* with higher tree turnover, a consequence of forest fragmentation *[@nascimento04a]* together with other global change drivers *[@laurance14]*, as well as may reflect possible differences in soil fertility *[@malhi04]*, all of which represent possible future research directions, beyond strictly neutral dynamics *[@terborgh95]*.
Furthermore, light availability at the plot level may not affect biomass storage, but instead individual light gaps may be more important for biomass dynamics *[@chazdon84]*.
    <!-- - although canopy light may still affect biomass generally in this case, without edge effects (check direct light-biomass correlation graph? -- no sig correlation) -->

Edge effects have been shown to lower both forest biomass and also diversity, and this study adds that the decline can be rapid and non-linear across edge gradients, along with supporting the likelihood of lasting changes to community composition.
The rapid diversity loss across the edge gradient may be in part due to finer variations in dispersal patterns among animal taxa, rather than overall dispersal mode more broadly.
Similarly, stochastic factors like priority effects after dispersal may also affect offspring establishment and re-growth.
While the biomass of the most abundant taxon *Vochysia* was resilient to edge effects, that of the key taxon for frugivores like birds and bats *Ficus* was significantly lowered across the forest edge, pointing to potentially different responses of biomass and food web associated ecosystem services.

<!-- # draft attic{-} -->

<!-- This study supports previous syntheses suggesting that forest biomass and taxonomic richness regenerate more quickly / are more resilient than diversity and community composition *[@poorter21?, other?]*. -->
<!-- This may be due to priority effects on successional trajectories acting on/at the stem/population level *[]*, variable turnover and reproduction rates among tree taxa *[]*, and/or differential dispersal and establishment success based on local [dispersive] animal community *[]*. ... -->
<!-- - mention more amundant genera patterns w/ *Vochysia* and *Ficus* as a key fruit-bearing taxon *[@tropicalbat/bird]*? -->

<!-- There is increasing interest in functional trait ecology of tropical trees *[@umaña##]*. -->
<!-- This focuses on a coarse functional association with successional stage, ... -->
<!-- but also includes patterns with wood density [underlying biomass estmations]... -->
<!-- - fig3 -- edge effect on succession generalists, but not specialists -->
<!-- The persistence of early-successional tree communities has been found to lower the standing biomass, likely due to lower wood density *[@césar21]*, and could therefore require a switch in focal timber species *[@karsten14]*. -->

<!-- # attic{-} -->

<!-- Under more sustainable forestry that maintains multiple functions, forest stand regeneration is likely more predictable, generally mirroring (i.e. near 1:1 replacement) the targeted tree cutting that is done using information collected about the ecological profiles and ages (when possible) of trees (Duah-Gyamfi et al 2014, Edwards et al 2014). -->

<!-- In contrast, under intensive silviculture that prioritizes the growth of select tree crop species, forest stands can show high wood production (Ferez et al 2015) while also usually showing negative trade-offs with other services, like that with biodiversity given reliance on monoculture planting designs. -->

<!-- In some cases where management decisions are made primarily based on regional economic but not local ecological information, due to limited access to funded scientific reporting (Cook et al 2012), forest stands can perform very poorly all-around. -->

<!-- After this more extreme type of management, it remains uncertain how natural regeneration in secondary forests will proceed, how local ecosystem services will be restored, and how restoration could be managed locally (Laurance et al 2012), e.g. based on surrounding forest proximity. -->

### 2022-07-11 article/merge.Rmd
start draft disc
first poster draft

---
title: ""
output: 
  html_notebook
  bookdown::html_document2
    number_sections: false
bibliography: ../../library.bib
---

# Abstract

```{r child='abstract.Rmd'}
```

# Intro

@@

# Refs

>>>>

---
title: "Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest"
author: 
  - Nicholas Medina #^a^*
  # - Max Villalobos? ^b^
  # - Andrea Johnson? ^b^
date: 
# scriptsize\
  Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI, USA 48109
  # ^a^ Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA, USA 02453; 
  # ^b^ Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica
# author: 
#   - name: "Nicholas Medina"
#     email: nmedina@umich.edu
#     affiliation: Brandeis University
#     orcid_id: 0000-0001-5465-3988
# address:
#   - code: Brandeis University
#     address: Biology and Environmental Studies, 415 South St, Waltham, MA 02453
# keywords: [1, 2, 3]

output: 
  # html_notebook
  # bookdown::html_document2:
  #   # df_print: paged
  #   global_numbering: TRUE
  #   toc: FALSE
  bookdown::pdf_document2:
    # number_sections: FALSE
    global_numbering: TRUE
    keep_tex: TRUE
    toc: FALSE
bibliography: 
  - ../../library.bib
  - ../../grateful-refs.bib
  - ../../software.bib
link-citations: TRUE
csl: ../../styles-master/elsevier-harvard.csl

fontsize: 12pt
geometry: margin=1in
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{setspace}\doublespace
# abstract
---

<!-- journals: J [Appl] Veg Sci, Basic Appl Ecol (3.5), Biotropica (2.5), Eco Evo -->

\
\

**Keywords**: secondary forest; regeneration; timber plantation associated biodiversity; edge effect; aboveground biomass; shade

\
\

***Corresponding author**: Nicholas Medina, nmedina@umich.edu

\newpage

# Abstract{-}

```{r child='abs.Rmd'}
```

\newpage

# Introduction

@@

\newpage

# Funding{-}

This study was funded by Brandeis University Hiatt Career Center World of Work Fellowship.

# Declaration of interests{-}

Authors declare no conflicts of interest.

# Acknowledgements{-}

Thanks to Professor Dan Perlman for initial networking; concurrent Osa Conservation managers Max Villalobos and Andrea Johnson for project guidance and supervision; Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance; concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support; peer James Fifer for discussion of early draft; and Luisa Valdez for translation assistance.

# Author contributions{-}

<!-- MV and AJ contributed to study conceptualization, administration, and supervision; NM and AJ contributed to funding acquisition; NM contributed to data collection; NM and MV contributed to data analysis;  -->
NM wrote initial draft and later revisions.

# Data statement{-}

Code stored at [github.com/nmedina17/osa](github.com/nmedin17/osa).

\newpage

# References
